Implementing Equational Constraints\\ in a Functional Language

### 2011-07-04 papers/wlp2011/paper.tex

Functional logic languages combine the most important
features of functional and logic programming in a single language
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they provide higher-order functions and demand-driven
evaluation from functional programming together with logic programming features
like non-deterministic search and computing with partial information
(logic variables).
This combination
has led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
and better abstractions for application programming, but it also gave rise
to new challenges of implementation.

@@

 \item compilation into non-strict functional languages like Haskell
and reusing the implementation
of lazy evaluation and higher-order functions
\cite{BrasselHuch07,BrasselHuch09}.
\end{enumerate}

The latter approach requires the implementation
of non-deterministic evaluations but has the advantage
that the explicit handling of non-determinism allows for
various search strategies like depth-first, breadth-first, parallel,
or iterative deepening instead of committing to a fixed (incomplete)
strategy like backtracking \cite{BrasselHuch07}.

In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselFischerHanusReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same results.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation in functional logic languages).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

@@

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are reducible to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e] =:= xs = e   where ys, e free
\end{curry}
%
Like in Haskell, most rules defining functions are \emph{constructor-based} 
\cite{ODonnell85}, i.e., in (\ref{rule})
$t_1,\ldots,t_n$ consist of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional patterns}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
defined operations. For instance, we can define the last element
of a list also by:
%
\begin{curry}
  last' (xs++[e]) = e
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last'} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns supports a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

Since all data types are extended by additional constructors,
we must also extend the definition of operations performing
pattern matching. For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is then extended as follows:
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
The second rule transforms a non-deterministic argument
into a non-deterministic result and
the final rule returns \code{Fail} in all other cases,
i.e., if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation (failure propagation).

@@

In Section \ref{sec:Unification} we stated that equational
constraints could also be implemented using pattern matching, but that
this implementation would increase the search space that has to be traversed 
to extract the values of a computation. To show the benificial effects
of our equality constraint implementations, we compare them to
the results using the equality operator \code{==} that
produces such a search space. The programs used for the benchmarks 
are partially taken from \cite{AntoyHanus05LOPSTR}.
%
\begin{quote}
\begin{description}
 \item[last]      Computes the last element of a list
 \item[simplify]  Simplification of symbolic arithmetic expressions
 \item[varInExp]  Nondeterministically returns a variable occuring
                  in a symbolic arithmetic expression
 \item[half]      Computes the half of a peano number using logic variables
 \item[palindrom] Checks if a list is a palindrom
%  \item[grep]
%  \item[horseman]
\end{description}
\end{quote}

Our benchmarks are executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor. Our compiler has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2). For later comparison with
other systems we considered PAKCS (Version 1.9.2, based on a SICStus-Prolog
4.1.2 backend) and MCC (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.

\begin{figure}
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Expression                              & \code{==} & \code{=:=} & \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 118.33    & 0.52       & 0.20        \\
% \code{grep}                             & 0.94      & 0.13       &             \\
\code{simplify}                         & 9.18      & 10.33      & 7.09        \\
\code{varInExp}                         & 0.50      & 86.27      & 0.14        \\
\code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
\code{palindrom}                        & 31.12     & 13.45      & 17.55       \\
% \code{horseman}                         & 3.88      & 4.93       &             \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different types of equation}
 \label{fig:search-strategies}
\end{figure}

\todo{Results from comparing, ==, =:=, =:<=}

As expected, the creation and traversal of a large search space introduced
by \code{==} in most cases is much slower than our presented approach for
unification. In addition, the \code{last} example shows that the lazy 
unification operator \code{=:<=} improves performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated. Interestingly, in the
\code{varInExp} example the strict unification performs significantly worse
than the other equality operators. \todo{Explain why}

In contrast to the Curry implementations PAKCS and MCC, out approach is based
upon an explicit representation of the search space and does not manipulate
a global state for unifying logic variables.
Nevertheless, our benchmarks show that it can in most cases
compete or even outperform these implementations.
Figure~\ref{fig:unification} shows a comparison of the curry systems where
strict equality (\ccode{=:=}) is used for unification.\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.} We can deduce from the results
that the unification of the MCC performs best. However, we can in most cases
outperform the Prolog based PAKCS but for some examples where further
research and optimization is necessary.

\begin{figure}
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Expression                              & KiCS2 & PAKCS & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.52  & 4.27   & 0.15 \\
% \code{grep}                             & 0.13  & 2.83   & 0.15 \\
\code{simplify}                         & 10.33 & 0.15   & 0.01 \\
\code{varInExp}                         & 86.27 & 0.35   & 0.03 \\
\code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
\code{palindrom}                        & 13.45 & 32.82  & 1.04 \\
% \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
\hline
\end{tabular}
\caption{Benchmarks: strict equational constraints}
 \label{fig:unification}
\end{figure}

As the MCC does not support functional patterns, the performance of lazy
unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
Again, our compiler performs well against PAKCS
and outperforms it in most cases. \todo{extend}

@@

Future work: adding other constraint systems

\bibliographystyle{plain}
\bibliography{mh}

>>>>

Functional logic languages combine the most important
features of functional and logic programming in a single language
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they provide higher-order functions and demand-driven
evaluation from functional programming together with logic programming features
like non-deterministic search and computing with partial information
(logic variables).
This combination
has led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
and better abstractions for application programming, but it also gave rise
to new implementation challenges.

@@

 \item compilation into non-strict functional languages like Haskell
and reusing the implementation
of lazy evaluation and higher-order functions
\cite{BrasselHuch07,BrasselHuch09}.
\end{enumerate}
%
The latter approach requires the implementation
of non-deterministic computations in a deterministic language
but has the advantage
that the explicit handling of non-determinism allows for
various search strategies like depth-first, breadth-first, parallel,
or iterative deepening instead of committing to a fixed (incomplete)
strategy like backtracking \cite{BrasselHuch07}.

In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselFischerHanusReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same results.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

@@

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are reducible to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e] =:= xs  = e   where ys, e free
\end{curry}
%
Like in Haskell, most rules defining functions are \emph{constructor-based} 
\cite{ODonnell85}, i.e., in (\ref{rule})
$t_1,\ldots,t_n$ consist of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional patterns}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
defined operations. For instance, we can define the last element
of a list also by:
%
\begin{curry}
  last' (xs++[e]) = e
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last'} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns support a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

Since all data types are extended by additional constructors,
we must also extend the definition of operations performing
pattern matching. For instance, consider the definition of polymorphic lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is then extended as follows:
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
The second rule transforms a non-deterministic argument
into a non-deterministic result and
the final rule returns \code{Fail} in all other cases,
i.e., if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation (failure propagation).

@@

In this section we evaluate our implementation by a set of benchmarks.
They were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2).
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a peano number using logic variables),
and
\code{palindrom} (check whether a list is a palindrom).

In Section \ref{sec:Unification} we discussed that equational
constraints could also be solved by generators without variable bindings,
but this technique might increase the search space due to the
possibly superfluous generation of all values.
To show the beneficial effects of our implementation
of equational constraints with variable bindings,
we compare the results of equational constraints to results
where the Boolean equality operator \code{==} is used (which
does not perform bindings but enumerate all values):

\begin{figure}
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=} & \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 118.33    & 0.52       & 0.20        \\
% \code{grep}                             & 0.94      & 0.13       &             \\
\code{simplify}                         & 9.18      & 10.33      & 7.09        \\
\code{varInExp}                         & 0.50      & 86.27      & 0.14        \\
\code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
\code{palindrom}                        & 31.12     & 13.45      & 17.55       \\
% \code{horseman}                         & 3.88      & 4.93       &             \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different types of equation}
 \label{fig:search-strategies}
\end{figure}

\noindent
As expected, the creation and traversal of a large search space introduced
by \code{==} is in most cases much slower than our presented approach
with variable bindings.
In addition, the example \code{last} shows that the lazy 
unification operator \code{=:<=} improves performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated. Interestingly, in the
\code{varInExp} example the strict unification performs significantly worse
than the other equality operators. \todo{Explain why}

In contrast to the Curry implementations PAKCS and MCC, our approach is based
on an explicit representation of the search space instead of manipulating
a global state containing bindings for logic variables.
Nevertheless, our benchmarks show that it can in most cases
compete with or even outperform these implementations.
Figure~\ref{fig:unification} shows a comparison of the Curry systems where
equational constraints are used in the examples.\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.} We can deduce from the results
that the unification of MCC performs best. However, in most cases
our implementation outperforms the Prolog-based PAKCS implementation,
except for some examples where further investigations
and optimizations are necessary.

\begin{figure}
\centering
\begin{tabular}{|l|l|l|l|}
\hline
Expression                              & KiCS2 & PAKCS & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.52  & 4.27   & 0.15 \\
% \code{grep}                             & 0.13  & 2.83   & 0.15 \\
\code{simplify}                         & 10.33 & 0.15   & 0.01 \\
\code{varInExp}                         & 86.27 & 0.35   & 0.03 \\
\code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
\code{palindrom}                        & 13.45 & 32.82  & 1.04 \\
% \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
\hline
\end{tabular}
\caption{Benchmarks: strict equational constraints}
 \label{fig:unification}
\end{figure}

\noindent
As MCC does not support functional patterns, the performance of lazy
unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
Again, our compiler performs well against PAKCS
and outperforms it in most cases. \todo{extend}

@@

We have presented an implementation of equational constraints
and functional patterns in a purely functional implementation
of Curry. Our implementation is based on adding binding constraints
to computed values and processing them when values are extracted
at the top level of a computation.
Since we have added only additional constructors and rules
for these constructors, our implementation does not cause
additional work if these features are not used, i.e.,
our implementation does not sacrifice the high efficiency
of the kernel implementation as described in
\cite{BrasselFischerHanusReck11}.
However, if these features are used, they usually lead
to a more efficient execution, as shown by our benchmarks.

Other implementations of equational constraints in functional logic
languages use side effects for their implementation.
For instance, PAKCS \cite{Hanus10PAKCS}
exploits the implementation of logic variables in Prolog,
which are implemented on the primitive level by side effects.
MCC \cite{Lux99FLOPS} compiles into C where a specific
abstract machine implements the handling of logic variables.
We have shown that our implementation is competitive to those.
In contrast to those systems, our implementation supports
a variety of search strategies like breadth-first or parallel search.

\todo{Future work: adding other constraint systems}

\bibliographystyle{mhnumber} %plain}
\bibliography{mh}
